---
title: "Germline Variant Calling — Accuracy Report"
subtitle: "Sample deliverable · HG002 chr20 benchmark"
author: "Peter Ristow, PhD — Principal Scientist (NGS)"
date: today
format:
  typst:
    papersize: a4
    margin:
      x: 1.8cm
      y: 1.8cm
    fontsize: 10pt
execute:
  echo: false
engine: knitr
---

```{r}
suppressPackageStartupMessages({library(dplyr); library(knitr); library(ggplot2); library(scales)})
m <- read.csv("results/benchmark_metrics.csv") |>
  mutate(caller = recode(caller, bcftools = "bcftools", gatk = "GATK"))
best <- m |> filter(var_type == "SNP") |> slice_max(f1, n = 1)
pct <- function(x) sprintf("%.2f%%", 100 * x)
```

**Executive summary.** Germline variants were called from the sequencing reads and **benchmarked
against the Genome-in-a-Bottle (GIAB) gold-standard truth set**. Two independent callers (bcftools
and GATK) were compared on identical data. Best SNP performance: **`r best$caller`** at
**F1 = `r sprintf('%.1f%%', 100*best$f1)`** (precision `r sprintf('%.1f%%', 100*best$precision)`,
recall `r sprintf('%.1f%%', 100*best$recall)`). This is an *objective* measure of correctness — not
an unverifiable VCF.

### Accuracy vs GIAB truth

```{r}
m |> arrange(var_type, caller) |>
  transmute(Caller = caller, Type = var_type,
            Precision = pct(precision), Recall = pct(recall), F1 = pct(f1),
            TP = tp, FP = fp, FN = fn) |>
  kable()
```

```{r}
#| fig-width: 5
#| fig-height: 2.6
ggplot(m, aes(recall, precision, colour = caller, shape = var_type)) +
  geom_point(size = 4) +
  scale_x_continuous(labels = percent) + scale_y_continuous(labels = percent) +
  labs(x = "Recall", y = "Precision", colour = NULL, shape = NULL) +
  theme_bw(base_size = 10)
```

**What this means.** SNP calls are recovered at high-99% precision and recall — the signature of a
correctly built short-read germline pipeline. Indels score lower (they are intrinsically harder),
which is why they are reported separately rather than hidden in a combined number.

**Caveats.** Metrics are computed inside GIAB high-confidence regions; difficult genomic regions
degrade any caller. For a sample without a truth set, accuracy is estimated against a matched
control or orthogonal data.

**Reproducibility.** The full pipeline (align → call → benchmark) re-runs from a clean clone in
minutes and is verified by continuous integration on every change. Code and live report:
`github.com/pgristow/variant-calling-giab-benchmark`.

---
*Prepared by Peter Ristow, PhD — Roche Principal Scientist (NGS). Production engagements use a
private, hardened pipeline on your data under NDA.*
